\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-12-15 sections/chameleon.tex
Some small fixes

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 35mm 0mm 0mm]{images/add-pair-example.pdf}
    \caption{
    A type error displayed in Visual Studio Code\cite{microsoft_visual_nodate} and the Haskell Vscode extension\cite{haskell_haskell_nodate}.
The expression \texttt{addPair} is blamed for causing the type error. This may not match the programmers' intention. 
    }
    \label{fig:motivation-example}
\end{figure}
\subsubsection{\textbf{Traditional type errors show only limited location}}
Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' The type error in Fig.~\ref{fig:motivation-example} can be fixed in multiple locations. For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}. In the type error message, only the \texttt{addPair} expression on line 4 was blamed.  In this small example, the whole context is visible, but the problem is worse for the programmer in large programs where the lines of code contributing to the type error are far apart in the source code.

\subsubsection{\textbf{Traditional type errors are biased}}
A common form of bias happens when a type error is reported in one expression, but it can occur in multiple other expressions as well. In Fig.~\ref{fig:motivation-example}, the error message arbitrarily focuses the users' attention on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect. %Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect. 
Another form of bias is that traditional type errors are often framed as conflicts between \texttt{Expected type} and \texttt{Actual type}. This framing is standard practice in most typed languages. However, what is the \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer. In both of these cases, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.

\subsubsection{\textbf{Traditional type errors give poor explanations}}
When the compiler rejects a program, the internal state of type checking is the result of complex computation. But the details of this process are hard to explain to users and are usually not reported by compilers. For the typical type error shown in Fig.~\ref{fig:motivation-example}, the evidence for the type error is gathered from the previous two declarations. These have to be rediscovered by programmers using less rigorous methods.

@@

\noindent\textbf{Do not presume} which expression is to blame for the type error based on the order of computation or which possible type for an expression is 'actual' or 'expected'

@@

The type compare tool shows conflicting types in different colors, each associated with one or more error locations highlighted in a matching color (Fig.~\ref{fig:compare}).  
%The type error must be fixed by modifying at least one of these locations. 
%Error locations are highlighted with the matching color of the conflicting types. The type compare tool is useful to quickly bisect the type error. 
If the programmers know the expression's intended type (they usually do), they will be able to eliminate half of the possible locations. 
%To make this bisecting effect more pronounced, a user can hover over one of the possible types and only show the relevant locations that contribute to that type. 
A hover interaction over one of the possible types facilitates such bisection, causing only the relevant locations that contribute to that type to be highlighted. 
%This is a convenient way to put the error \emph{under the spotlight}.

@@

Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are well-typed. Once a card is selected, the information in the conflicting types block changes to reflect the change of candidate expression. In the editor pane, some error locations change highlight colors based on the updated candidate expression.

This feature can be effective in resolving the problem that traditional tools tend to report errors that happened in source code from third-party libraries. With candidate expression cards, a programmer can freely change the context of the type error to the user-defined variables and functions to gain a better understanding of why their own code does not match the library code instead of the other direction.

Programmers select a candidate expression card by clicking on one card. Alternatively, they can peek at the effect of the change by hovering on one card. The hover effect is reverted once the cursor moves away.

@@

Deduction steps allow programmers to explore all the error locations one at a time (Fig. \ref{fig:deduction}). Steps are shown as a list of sequentially  numbered circular buttons (step buttons) and an explanation layer in the editor window. In the explanation layer, two highlighted error locations are outlined, and a line is drawn to connect these two locations. This line is accompanied by a human-readable text explanation of their semantic connection. Programmers are free to activate any step. The active step is shown as the green button. When activating a step, some highlights switch color. The message in the explanation layer changes as well. A program in Fig. \ref{fig:deduction} generates a list of steps shown in Fig. \ref{fig:step-interface} left.

Programmers can use mouse and keyboard shortcuts to increment or decrement the step number or jump to any step. Typically, programmers resolve  type errors by navigating through all the deduction steps and verifying whether each explanation aligns with the intended meaning of the program. Eventually, they will find a step that does not match their intention, and the type error can be fixed by modifying one of the two outlined error locations.

@@

Chameleon was originally a command-line tool developed in the early 2000s to improve type error reporting %and introduced a design for interactive type debugging 
for the Haskell programming language.
Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ben_gamari_home_2022}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving. In Chameleon, Constraints are generated from the source code based on typing rules. In addition, each constraint is labeled with the location (line and column numbers) where it is generated. This set of constraints is consistent if the program is well-typed and inconsistent otherwise. When a type error occurs, an efficient algorithm is used to derive a minimal subset of the constraints that still contain inconsistencies. This subset is called a Minimal Unsatisfiable Subset (MUS). From this, Chameleon can report a list of locations, using the labels of constraints that are in the MUS. Stuckey et al.~\cite{stuckey_interactive_2003} showed that program locations linked to the constraints from an MUS are all relevant to the type error and must include the cause of the error.

@@

Using concrete types provides extra information to programmers. With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space. However, this information is not preserved in \texttt{(a, Float)}. If the types get too verbose for programmers to identify the conflict between the two signatures, it is always possible to mute the overlapping information and highlight the differences from the debugging front end.

@@

In addition, \chameleon{} provides support for type explanation. Similar to the type explanation system  in ~\cite{jun_explaining_2002},  \chameleon{} is able to produce a human-readable explanation, but for type errors. This is achieved by annotating nodes in the abstract syntax tree with constraints and the type of rules applied. We generate an inference history from constraints and accompanying annotations.

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 35mm 0mm 0mm]{images/add-pair-example.pdf}
    \caption{
    A type error displayed in Visual Studio Code\cite{microsoft_visual_nodate} and the Haskell Vscode extension\cite{haskell_haskell_nodate}.
The expression \texttt{addPair} is blamed for causing the type error. This may not match the programmers' intention. 
    }
    \label{fig:motivation-example}
\end{figure}
\subsubsection{\textbf{Traditional type errors show only limited location}}
Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' The type error in Fig.~\ref{fig:motivation-example} can be fixed in multiple locations. For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}. In the type error message, only the \texttt{addPair} expression on line 4 was blamed.  In this small example, the whole context is visible, but it can be more problematic in large programs where the lines contributing to the type error are far apart in the source code.

\subsubsection{\textbf{Traditional type errors are biased}}
A common form of bias happens when a type error is reported in one expression, but it can occur in multiple other expressions as well. In Fig.~\ref{fig:motivation-example}, the error message arbitrarily focuses on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect. %Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect. 
Another form of bias is that traditional type errors are often framed as conflicts between \texttt{Expected type} and \texttt{Actual type}. This framing is standard practice in most typed languages. However, what is the \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer. In both forms, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.

\subsubsection{\textbf{Traditional type errors give poor explanations}}
When the compiler rejects a program, the internal state of type checking is the result of a complex computation. But the details of this process are hard to explain to users and are usually not reported by compilers. For the typical type error shown in Fig.~\ref{fig:motivation-example}, the evidence for the type error is gathered from the previous two declarations. These have to be rediscovered by programmers using less rigorous methods.

@@

\noindent\textbf{Do not presume} which expression is to blame for the type error based on the order of computation or which possible type for an expression is 'actual' or 'expected'.

@@

The type compare tool shows conflicting types in different colors, each type associated with one or more error locations highlighted in a matching color (Fig.~\ref{fig:compare}).  
%The type error must be fixed by modifying at least one of these locations. 
%Error locations are highlighted with the matching color of the conflicting types. The type compare tool is useful to quickly bisect the type error. 
If the programmers know the expression's intended type (they usually do), they will be able to eliminate half of the possible locations. 
%To make this bisecting effect more pronounced, a user can hover over one of the possible types and only show the relevant locations that contribute to that type. 
A hover interaction over one of the possible types facilitates such bisection, causing only the relevant locations that contribute to that type to be highlighted. 
%This is a convenient way to put the error \emph{under the spotlight}.

@@

% This feature can be effective in resolving the problem that traditional tools tend to report errors that happened in source code from third-party libraries. With candidate expression cards, a programmer can freely change the context of the type error to the user-defined variables and functions to gain a better understanding of why their own code does not match the library code instead of the other direction.

Programmers select a candidate expression card by clicking on one card. Once a card is selected, the information in the conflicting types block changes to reflect the change of candidate expression. In the editor pane, some error locations change highlight colors based on the updated candidate expression. Alternatively, programmers can preview the change of a candidate expression by hovering on one card. The hover effect is reverted once the cursor moves away.

@@

Deduction steps allow programmers to explore all the error locations one at a time (Fig. \ref{fig:deduction}). Steps are shown as a list of sequentially  numbered circular buttons (step buttons) and an explanation layer in the editor window. In the explanation layer, the two locations under examination are outlined, and a line is drawn to connect these two locations. This line is accompanied by a human-readable text explanation of their semantic connection. Programmers are free to activate any step. The active step is shown in green. When activating a step, some highlights switch color. The message in the explanation layer changes as well. A program in Fig. \ref{fig:deduction} generates a list of steps shown in Fig. \ref{fig:step-interface} left.

Programmers can use mouse and keyboard shortcuts to increment or decrement the step number or jump to any step. Programmers resolve type errors by navigating through all the deduction steps and verifying whether each explanation aligns with their intention. Eventually, they will find a step that does not match, and the type error can be fixed by modifying one of the two outlined locations.

@@

Chameleon was originally a command-line tool developed in the early 2000s to improve type error reporting %and introduced a design for interactive type debugging 
for the Haskell programming language.
Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ben_gamari_home_2022}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving. In Chameleon, constraints are generated from the source code based on typing rules. In addition, each constraint is labeled with the location where it is generated. This set of constraints is consistent if the program is well-typed and inconsistent otherwise. When a type error occurs, an efficient algorithm is used to derive a minimal subset of the constraints that still contain inconsistencies. This subset is called a Minimal Unsatisfiable Subset (MUS). From this, Chameleon can report a list of locations, using the labels of constraints that are in the MUS. Stuckey et al.~\cite{stuckey_interactive_2003} showed that program locations linked to the constraints from an MUS are all relevant to the type error and must include the cause of the error.

@@

Using concrete types provides extra information to programmers. With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space. However, a type of \texttt{(a, Float)} does not preserve such information. If the types get too verbose for programmers to identify the conflict between the two signatures, it is always possible to mute the overlapping information and highlight the differences from the debugging front end.

@@

In addition, \chameleon{} provides support for type explanation. Similar to the type explanation system  in ~\cite{jun_explaining_2002},  \chameleon{} is able to produce a human-readable explanation, but for type errors. This is achieved by annotating nodes in the abstract syntax tree with constraints and the typing rules used. We generate an inference history from constraints and accompanying annotations.

### 2022-12-15 sections/conclusion.tex
Some small fixes

We present \chameleon{}, a type debugging tool for the Haskell programming language. Its constraint-based type inference engine provides unbiased and comprehensive error location reporting. 
% This sentence could go away:
%\chameleon{} also features an interactive debugging interface with three debugging modes, enabling programmers to interactively explore the stepwise explanations. 
Our studies evaluated the tool's design with programmers. We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages. Further, programmers actively using \chameleon{} interactive features are shown to fix type errors faster than simply reading the type error output.
\chameleon{} currently works with the Haskell language, but in future we plan to extend the typechecking system to work with other strongly typed languages, such as Rust or TypeScript.

>>>>

We present \chameleon{}, a type debugging tool for the Haskell programming language. Its constraint-based type inference engine provides unbiased and comprehensive error location reporting. 
% This sentence could go away:
%\chameleon{} also features an interactive debugging interface with three debugging modes, enabling programmers to interactively explore the stepwise explanations. 
Our studies evaluated the tool's design with programmers. We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages. Further, programmers actively using \chameleon{} interactive features are shown to fix type errors faster than simply reading the type error output.
\chameleon{} currently works with the Haskell language, but in the future, we plan to extend the type-checking system to work with other strongly typed languages, such as Rust or TypeScript.

### 2022-12-15 sections/discussion.tex
Some small fixes

\subsection{Externalise Intermediate Typing Inforamtion}
We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information. With the program shown in Listing~\ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}. Although, in theory, these two statements have the same effect, programmers are often required to compute the latter from the former or vice versa. And this computation may carry out multiple layers. Programmers have to remember all the intermediate types and their reasoning throughout such mental gymnastics. Assisted by candidate expression cards and deduction steps, this intermediate information is externalized on screen and can be retrieved anytime. A recent study on working memory \cite{crichton_role_2021} suggested this approach may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.

>>>>

\subsection{Externalize Intermediate Typing Information}
We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information. With the program shown in Listing~\ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}. Although  these two statements are equivalent in theory, programmers are often required to compute the latter from the former or vice versa. And this computation may carry out multiple layers. Programmers have to remember all the intermediate types and their reasoning throughout such mental gymnastics. Assisted by candidate expression cards and deduction steps, this intermediate information is externalized on screen and can be retrieved anytime. A recent study on working memory \cite{crichton_role_2021} suggested this approach may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.

### 2022-12-15 sections/evaluation.tex
Updated bib
Some small fixes

Participants were recruited via the Reddit \textit{r/haskell} and \textit{r/programminglanguages} communities. 
%Recruiting from social media allowed us to access a more diverse demographic that better represent the true population of Haskell programmers. 
Participation is fully anonymized; detailed ethical implications of these experiments are reviewed and approved by the IRB of the institution of the authors.

\todo{Need more detail on participants - years experience, male vs female, etc}

@@

We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions. This provides us with a "poor man's video recording" to identify usability issues in the study and to recognize general patterns. 
%The reason we found this format inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.
%One use of the session recording is for outlier trimming.  We identify participants who left the computer for an extended period of time. We first ran a statistical analysis to identify the long gap (2 standard deviations) between actions using program logs for each participant. Then, we manually checked with the session recording to verify there was actually no user input (mouse movement, scrolling).

@@

Eight tasks were given in both studies. In study 1a, the tasks were taken from the exercises of the Haskell programming class in the authors' institute. These tasks cover a variety of type errors.  In the second study, the tasks are sourced from the top 20 Haskell topics on GitHub~\cite{github_github_2022}. The authors then manually added type errors into the program. In both studies, the type errors include simple mismatch, confusing syntax, missing instance, precedence and fixation, infinite types, and confusing list versus element. The categories of type errors to include in the tasks followed the common type errors in Tirronen's study \cite{tirronen_understanding_2015}.

@@

The data collected during study 1a, Fig.~\ref{fig:analysis-1a} does not show significant differences across Tasks 1-7. In hindsight, these tasks were trivial challenges for most users and the individual differences among participants are generally more significant than the differences between treatments. However, one interesting observation is task 8, where the \chameleon{} group outperformed the GHC group (p-value = 0.002456, Wilcoxin signed test). We attribute this significant difference to the hardness of Task 8. The source file is longer and involves more language features (abstract data types and high-level functions). GHC struggles to produce a relevant error message for this type of error. 
From this result, we hypothesized that we may observe a more significant difference using tasks with lengthier and more realistic source code. 
This hypothesis is also supported by the most common feedback (7 out of 35) that the tasks were too trivial to invite meaningful evaluation. One participant said, ``Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."

Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group was faster (p = 0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1. For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity. For all other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program. In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. ``It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. ``I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgments. This makes it quite easy to figure out how to rectify errors." reported another participant.

@@

\subsubsection{\textbf{\chameleon{} 2}}  \label{sub:us4}
Based on observations of Study 1 we introduced several new features to \chameleon{}, eventually resulting in the UI depicted in Figs.~(2-13). Advanced debugging features such as deduction steps and candidate expressions were available in this iteration. In addition, mode switching is enabled by default. A few other user interfaces \cite{anonymous_interactive_2021} were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2. Study 2 therefore addresses the research question:

@@

During the study, the initial mode of each task alternated through the three different modes and repeated three cycles in nine tasks. The order of the three modes in each cycle is counterbalanced among all participants. However, participants can switch to other modes at any time.

@@

Study 2 is more exploratory in methodology than Study 1. 
We were expecting to let programmers discover their own way of using the tool. In post hoc analysis of the collected log data, we were able to extrapolate some interesting patterns of how the tool was used. %We share a few of our observations in this section.

@@

As shown in  Fig.~\ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better, and the lowest interaction level was worse (one-way ANOVA since we have three samples, p = 0.0368). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As with Study 1a and 1b, this difference is likely related to task difficulty. Tasks 4 and 6 are shorter than other tasks. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). These may be useful because following the natural reading order will allow participants to skip lines on the bottom, and thus the advantage of \chameleon{} is less critical. On the other hand, task 9 is the lengthiest task of all. It also involves more complex features, such as mutually recursive type definitions.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 15mm 0mm 35mm,clip]{images/user-study-2.pdf}
    \caption{Study 2 task completion time (secs.) with 95\% confidence intervals.}
    \label{fig:r4-analysis}
\end{figure}

@@

Another observation is when using the mode switching feature of \chameleon{}, we find presenting the starting mode and finishing mode of each task and each participant in a correlation matrix is very telling (Fig. \ref{fig:r4-mode-switching}). This observation suggests two characteristics of using multi-mode debugging tools. First, to answer \textbf{RQ2.2} programmers are generally reluctant to switch modes (146/277 changing modes vs. 131/277 staying in the same mode). Second, to answer \textbf{RQ2.3} when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
\begin{figure}
    \centering
    \includegraphics[width=0.7\linewidth,trim=0mm 5mm 0mm 5mm,clip]{images/mode-switching.pdf}
    \caption{Study 2 mode switches by starting mode.  Users overwhelmingly switched to the more sophisticated interface mode.
        % The rows represent the finishing mode, columns starting modes.
    }
    \label{fig:r4-mode-switching}
\end{figure}

@@

One threat to the validity of the evaluation is the number of participants. Although for each study we received hundreds of online participants, the studies suffer from a high abandonment rate (especially study 1b). This was expected: the programming challenges are difficult and our volunteer participants are unremunerated. 
%To encourage completion, we included a cheat sheet containing a summary of the key functionality of the tools and necessary type signatures. A skip button is available for when participants are stuck.
Because we recruited participants online and anonymized all the participants, it is possible for participants of a previous study to enter a later one. This creates variation in familiarity. We offset this by using new code challenges in every study and conducting trial runs before data collection to bring new participants up to speed.
Conducting studies remotely and unsupervised there is no way to intervene when users encounter usability issues. To mitigate this we conducted cognitive walkthroughs and sandbox pilots before running each study.

Future evaluation would benefit from using more realistic tasks. In studies 1b and 2, we used more challenging tasks compared to study 1a. However, the complexity is just short of what professional Haskell programmers may face in the production Haskell codebase. It would be interesting to see how \chameleon{} is used against type errors that span multiple files and packages and include confusing abstractions, like Monads, Monad transformers, and Lenses.

>>>>

Participants were recruited via the Reddit \textit{r/haskell} and \textit{r/programminglanguages} communities. 
%Recruiting from social media allowed us to access a more diverse demographic that better represent the true population of Haskell programmers. 
Participation is fully anonymized; detailed ethical implications of these experiments are reviewed and approved by the IRB of the authors' institution.

@@

We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions. This provides us with a ``poor man's video recording" to identify usability issues in the study and to recognize general patterns. 
%The reason we found this format inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.
%One use of the session recording is for outlier trimming.  We identify participants who left the computer for an extended period of time. We first ran a statistical analysis to identify the long gap (2 standard deviations) between actions using program logs for each participant. Then, we manually checked with the session recording to verify there was actually no user input (mouse movement, scrolling).

@@

Eight tasks were given in both studies. In study 1a, the tasks were taken from the exercises of the Haskell programming class in the authors' institute. These tasks cover a variety of type errors.  In the second study, the tasks are sourced from the top 20 Haskell topics on GitHub~\cite{github_github_2022}. The authors then manually added type errors into the program. In both studies, the type errors include simple mismatch, confusing syntax, missing instance, precedence and fixation, infinite types, and confusing list versus element. These categories follow the common type errors in Tirronen's study \cite{tirronen_understanding_2015}.

@@

The data collected during study 1a, Fig.~\ref{fig:analysis-1a} does not show significant differences across Tasks 1-7. In hindsight, these tasks were trivial challenges for most users, and the individual differences among participants are generally more significant than the differences between treatments. However, one interesting observation is task 8, where the \chameleon{} group outperformed the GHC group. We attribute this significant difference to the difficulty of Task 8. The source file is longer and involves more language features (abstract data types and high-level functions). GHC struggles to produce a relevant error message for this type of error. 
From this result, we hypothesized that we might observe a more significant difference using tasks with lengthier and more realistic source code. 
This hypothesis is also supported by the most common feedback (7 out of 35) that the tasks were too trivial to invite meaningful evaluation. One participant said, ``Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."

Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group was faster than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1. A two-sample paired t-test was performed to compare the completion time between \chameleon{} and GHC groups. There was a significant difference between the two groups: $t(23) = -3.86, p = 0007$. For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity. For all other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program. In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. ``It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. ``I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgments. This makes it quite easy to figure out how to rectify errors." reported another participant.

@@

\subsubsection{\textbf{\chameleon{} 2}}  \label{sub:us4}
Based on observations of Study 1 we introduced several new features to \chameleon{}, eventually resulting in the UI depicted in Figs.~(2-13). Interactive features were available in this iteration, such as deduction steps, candidate expressions, and mode switching. A few other user interfaces \cite{anonymous_interactive_2021} were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2. Study 2 addresses the research question:

@@

During each run, the initial mode of each task alternated through the three different modes and repeated three cycles in nine tasks. The order of the three modes in each cycle is counterbalanced among all participants. However, participants can switch to other modes at any time.

@@

Study 2 is more exploratory in methodology than Study 1. We encouraged programmers  to discover their way of using the tool. In post hoc analysis of the collected log data, we were able to extrapolate some interesting patterns of how the tool was used.

@@

As shown in  Fig.~\ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better, and the lowest interaction level was worse. Tukey’s HSD Test for multiple comparisons found that the completion time was significantly different between the minimal interaction group and the high interaction group ($p \le 0.001$, 95\% C.I. = [18.26, 31.41]), and between the minimal interaction group and the low interaction group ($p \le 0.001$, 95\% C.I. = [11.96, 26.67]). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As with Study 1a and 1b, this difference is likely related to task difficulty. Tasks 4 and 6 are shorter than other tasks. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). Users simply reading top to bottom could quickly identify the error without needing to skip unrelated sections of code using the information provided by /chameleon{}. This reduced the apparent benefit of /chameleon{} in these tasks. On the other hand, task 9 is the lengthiest task of all. It also involves more complex features, such as mutually recursive type definitions.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 15mm 0mm 50mm,clip]{images/user-study-2.pdf}
    \caption{Study 2 task completion time (secs.) with 95\% confidence intervals.}
    \label{fig:r4-analysis}
\end{figure}

@@

Another observation is when using the mode switching feature of \chameleon{}, we show this by presenting the starting mode and finishing mode of each task and each participant in a correlation matrix (Fig. \ref{fig:r4-mode-switching}). This observation suggests two characteristics of using multi-mode debugging tools. First, to answer \textbf{RQ2.2} programmers are generally reluctant to switch modes: 53\% changing modes vs. 47\% staying in the same mode. Second, to answer \textbf{RQ2.3} when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
\begin{figure}
    \centering
    \includegraphics[width=0.7\linewidth,trim=0mm 5mm 0mm 5mm,clip]{images/mode-switching.pdf}
    \caption{Study 2 mode switches by starting mode.  Users overwhelmingly switched to the more sophisticated interface mode.
        % The rows represent the finishing mode, columns starting modes.
    }
    \label{fig:r4-mode-switching}
\end{figure}

@@

One threat to the validity of the evaluation is the number of participants. Although for each study we received hundreds of online participants, the studies suffered from a high abandonment rate (especially study 1b). This was expected: the programming challenges are difficult, and our volunteer participants are unremunerated. 
Because we recruited participants online and anonymized all the participants, it is possible for participants of a previous study to enter a later one. This creates variation in familiarity. We offset this by using new code challenges in every study and conducting trial runs before data collection to bring new participants up to speed.
Conducting studies remotely and unsupervised there is no way to intervene when users encounter usability issues. To mitigate this, we conducted cognitive walkthroughs and sandbox pilots before running each study.

Future evaluation would benefit from using more realistic tasks. In studies 1b and 2, we used more challenging tasks compared to study 1a. However, the complexity is just short of what professional Haskell programmers may face in the production Haskell codebase. It would be interesting to see how \chameleon{} is used against type errors that span multiple files and packages and include more confusing abstractions, like Monads, Monad transformers, and Lenses.
